Taking Linear Logic Apart

### 2018-04-26 doc/tlla2018/main.tex
Finished section on CP; renamed preservation to subject reduction

We proceed as follows.
We introduce \cp and restate its proofs of subject reduction and progress
(\cref{sec:cp}), then introduce \hcp and give its proofs of subject reduction
and progress (\cref{sec:hcp}).
We then prove several properties relating the two. Specifically, we prove that
every \cp process is an \hcp process, that every \hcp process corresponds to
several entangled \cp processes, and that \hcp supports the same communication
protocols as \cp, and no more (\cref{sec:cp2hcp}).
Finally, we discuss related work and conclude
(\cref{sec:related-work,sec:conclusions}).

@@

The term language of \cp is a variant of the \textpi-calculus.
\begin{definition}[Terms]\label{def:cp-terms}
  \[
    \begin{array}{rllrll}
      \tm{P}, \tm{Q}, \tm{R}
           :=& \tm{\cpLink{x}{y}}       &\text{link}
      &  \mid& \tm{\cpCut{x}{P}{Q}}     &\text{parallel composition, ``cut''}
      \\ \mid& \tm{\cpSend{x}{y}{P}{Q}} &\text{output}
      &  \mid& \tm{\cpRecv{x}{y}{P}}    &\text{input}
      \\ \mid& \tm{\cpHalt{x}}          &\text{halt}
      &  \mid& \tm{\cpWait{x}{P}}       &\text{wait}
      \\ \mid& \tm{\cpInl{x}{P}}        &\text{select left choice}
      &  \mid& \tm{\cpInr{x}{P}}        &\text{select right choice}
      \\ \mid& \tm{\cpCase{x}{P}{Q}}    &\text{offer binary choice}
      &  \mid& \tm{\cpAbsurd{x}}        &\text{offer nullary choice}
    \end{array}
  \]
\end{definition}
The variables $\tm{x}$, $\tm{y}$, and $\tm{z}$ range over channel names.
The construct $\tm{\cpLink{x}{y}}$ links two
channels~\cite{sangiorgi1996,boreale1998}, forwarding messages received on
$\tm{x}$ to $\tm{y}$ and vice versa.
The construct $\tm{\cpCut{x}{P}{Q}}$ creates a new channel $\tm{x}$, and composes
two processes, $\tm{P}$ and $\tm{Q}$, which communicate on $\tm{x}$, in parallel.
Therefore, in $\tm{\cpCut{x}{P}{Q}}$ the name $\tm{x}$ is bound in both $\tm{P}$
and $\tm{Q}$.
In $\tm{\cpRecv{x}{y}{P}}$ and $\tm{\cpSend{x}{y}{P}{Q}}$, round brackets denote
input, square brackets denote output.
We use bound output~\cite{sangiorgi1996}, meaning that both input and output
bind a new name.
In $\tm{\cpRecv{x}{y}{P}}$ the new name $\tm{y}$ is bound in $\tm{P}$.
In $\tm{\cpSend{x}{y}{P}{Q}}$, the new name $\tm{y}$ is only bound in $\tm{P}$,
while $\tm{x}$ is only bound in $\tm{Q}$.

@@

\begin{definition}[Reduction]\label{def:cp-reduction}
  A reduction $\reducesto{P}{Q}$ denotes that the process $\tm{P}$ can reduce
  to the process $\tm{Q}$ in a single step.
  The relation $\Longrightarrow^\star$ is the reflexive, transitive closure of
  $\Longrightarrow$.
  Reductions can only be constructed as follows:
  \[
    \begin{array}{llcll}
      \cpRedAxCut1
      & \tm{\cpCut{x}{\cpLink{w}{x}}{P}}
      & \Longrightarrow
      & \tm{\cpSub{w}{x}{P}}
      \\
      \cpRedBetaTensParr
      & \tm{\cpCut{x}{\cpSend{x}{y}{P}{Q}}{\cpRecv{x}{y}{R}}}
      & \Longrightarrow
      & \tm{\cpCut{y}{P}{\cpCut{x}{Q}{R}}}
      \\
      \cpRedBetaOneBot
      & \tm{\cpCut{x}{\cpHalt{x}}{\cpWait{x}{P}}}
      & \Longrightarrow
      & \tm{P}
      \\
      \cpRedBetaPlusWith1
      & \tm{\cpCut{x}{\cpInl{x}{P}}{\cpCase{x}{Q}{R}}}
      & \Longrightarrow
      & \tm{\cpCut{x}{P}{Q}}
      \\
      \cpRedBetaPlusWith2
      & \tm{\cpCut{x}{\cpInr{x}{P}}{\cpCase{x}{Q}{R}}}
      & \Longrightarrow
      & \tm{\cpCut{x}{P}{R}}
      \\
      \cpRedKappaTens1
      & \tm{\cpCut{x}{\cpSend{y}{z}{P}{Q}}{R}}
      & \Longrightarrow \;
      & \tm{\cpSend{y}{z}{\cpCut{x}{P}{R}}{Q}}
        \quad \text{if} \; \notFreeIn{x}{Q}
      \\
      \cpRedKappaTens2
      & \tm{\cpCut{x}{\cpSend{y}{z}{P}{Q}}{R}}
      & \Longrightarrow \;
      & \tm{\cpSend{y}{z}{P}{\cpCut{x}{Q}{R}}}
        \quad \text{if} \; \notFreeIn{x}{P}
      \\
      \cpRedKappaParr
      & \tm{\cpCut{x}{\cpRecv{y}{z}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpRecv{y}{z}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaBot
      & \tm{\cpCut{x}{\cpWait{y}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpWait{y}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaPlus1
      & \tm{\cpCut{x}{\cpInl{y}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpInl{y}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaPlus2
      &\tm{\cpCut{x}{\cpInr{y}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpInr{y}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaWith
      & \tm{\cpCut{x}{\cpCase{y}{P}{Q}}{R}}
      & \Longrightarrow \;
      & \tm{\cpCase{y}{\cpCut{x}{P}{R}}{\cpCut{x}{Q}{R}}}
      \\
      \cpRedKappaTop
      & \tm{\cpCut{x}{\cpAbsurd{y}}{R}}
      & \Longrightarrow \;
      & \tm{\cpAbsurd{y}}
    \end{array}
  \]
  \begin{center}
    \begin{prooftree*}
      \AXC{$\reducesto{P}{P^\prime}$}
      \SYM{\cpRedGammaCut}
      \UIC{$\reducesto{\cpCut{x}{P}{Q}}{\cpCut{x}{P^\prime}{Q}}$}
    \end{prooftree*}
    \begin{prooftree*}
      \AXC{$\tm{P}\equiv\tm{Q}$}
      \AXC{$\reducesto{Q}{Q^\prime}$}
      \AXC{$\tm{Q^\prime}\equiv\tm{P^\prime}$}
      \SYM{\cpRedGammaEquiv}
      \TIC{$\reducesto{P}{P^\prime}$}
    \end{prooftree*}
  \end{center}
\end{definition}

\begin{definition}[Types]\label{def:cp-types}
  \[
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
    \end{array}
  \]
\end{definition}

\begin{figure*}[!htb]
  Structural Rules.
  \begin{center} \cpInfAx     \cpInfCut      \end{center}

  Logical Rules.
  \begin{center} \cpInfTens   \cpInfParr     \end{center}
  \begin{center} \cpInfOne    \cpInfBot      \end{center}
  \begin{center} \cpInfPlus1  \cpInfPlus2    \end{center}
  \begin{center} \cpInfWith                  \end{center}
  \begin{center} \cpInfNil    \cpInfTop      \end{center}

  \caption{Classical Processes (\cp)}
  \label{fig:cp}
\end{figure*}

\begin{lemma}[Preservation for $\equiv$]\label{lem:cp-preservation-equiv}
  If $\tm{P}\equiv\tm{Q}$, then $\seq[P]{\Gamma}$ iff $\seq[Q]{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\tm{P}\equiv\tm{Q}$.
\end{proof}

\begin{theorem}[Preservation]\label{thm:cp-preservation}
  If $\seq[P]{\Gamma}$ and $\reducesto{P}{Q}$, then $\seq[Q]{\Gamma}$.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reducesto{P}{Q}$.
\end{proof}

\begin{theorem}[Progress]\label{thm:cp-progress}
  If $\seq[P]{\Gamma}$, then there exists a $\tm{Q}$ such that
  $\tm{P}\Longrightarrow^\star\tm{Q}$ and $\tm{Q}$ is not a cut.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\seq[P]{\Gamma}$.
  If the last rule is \textsc{Cut}, there are four cases:
  \begin{enumerate*}[label={\alph*)}]
  \item
    if either side of the cut is an axiom, we apply \cpRedAxCut1;
  \item
    if either side of the cut is itself a cut, we recursively eliminate the cut;
  \item
    if both sides are logical rules acting on the cut formula, we apply
    one of the $\beta$-rules;
  \item
    otherwise, at least one side is a logical rule acting on a formula other
    than the cut formula, in which case we apply one of the $\kappa$-rules.
  \end{enumerate*}
\end{proof}

@@

\begin{lemma}[Preservation for $\equiv$]\label{lem:hcp-preservation-equiv}
  If $\tm{P}\equiv\tm{Q}$, then $\seq[P]{\mathcal{G}}$ iff $\seq[Q]{\mathcal G}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\tm{P}\equiv\tm{Q}$.
\end{proof}

\begin{theorem}[Preservation]\label{thm:hcp-preservation}
  If $\seq[P]{\mathcal{G}}$ and $\reducesto{P}{Q}$, then $\seq[Q]{\mathcal G}$.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reducesto{P}{Q}$.
\end{proof}

>>>>

In this paper, we introduce the Hypersequent Classical Processes (\hcp), which
addresses this mismatch.
The key insight is to register parallelism in the typing judgements using
hypersequents, a technique from logic which generalises judgements from one
sequent to many.
This allows us to take apart the term constructs used in Classical Processes
(\cp) to more closely match those of the standard $\pi$-calculus.
We proceed as follows.
We start by introducing \cp (\cref{sec:cp}).
Then, we introduce \hcp and prove it enjoys subject reduction and progress
(\cref{sec:hcp}).
We prove several properties relating the two.
Specifically, we prove that every \cp process is an \hcp process, that every
\hcp process corresponds to several entangled \cp processes, and that \hcp
supports the same communication protocols as \cp, and no more
(\cref{sec:cp2hcp}).
Finally, we discuss related work and conclude
(\cref{sec:related-work,sec:conclusions}).

@@

The term language of \cp is a variant of the \textpi-calculus.
The variables $\tm{x}$, $\tm{y}$, and $\tm{z}$ range over channel names.
The construct $\tm{\cpLink{x}{y}}$ links two
channels~\cite{sangiorgi1996,boreale1998}, forwarding messages received on
$\tm{x}$ to $\tm{y}$ and vice versa.
The construct $\tm{\cpCut{x}{P}{Q}}$ creates a new channel $\tm{x}$, and composes
two processes, $\tm{P}$ and $\tm{Q}$, which communicate on $\tm{x}$, in parallel.
Therefore, in $\tm{\cpCut{x}{P}{Q}}$ the name $\tm{x}$ is bound in both $\tm{P}$
and $\tm{Q}$.
In $\tm{\cpRecv{x}{y}{P}}$ and $\tm{\cpSend{x}{y}{P}{Q}}$, round brackets denote
input, square brackets denote output.
We use bound output~\cite{sangiorgi1996}, meaning that both input and output
bind a new name.
In $\tm{\cpRecv{x}{y}{P}}$ the new name $\tm{y}$ is bound in $\tm{P}$.
In $\tm{\cpSend{x}{y}{P}{Q}}$, the new name $\tm{y}$ is only bound in $\tm{P}$,
while $\tm{x}$ is only bound in $\tm{Q}$.
\begin{definition}[Terms]\label{def:cp-terms}
  \[
    \begin{array}{rllrll}
      \tm{P}, \tm{Q}, \tm{R}
           :=& \tm{\cpLink{x}{y}}       &\text{link}
      &  \mid& \tm{\cpCut{x}{P}{Q}}     &\text{parallel composition, ``cut''}
      \\ \mid& \tm{\cpSend{x}{y}{P}{Q}} &\text{output}
      &  \mid& \tm{\cpRecv{x}{y}{P}}    &\text{input}
      \\ \mid& \tm{\cpHalt{x}}          &\text{halt}
      &  \mid& \tm{\cpWait{x}{P}}       &\text{wait}
      \\ \mid& \tm{\cpInl{x}{P}}        &\text{select left choice}
      &  \mid& \tm{\cpInr{x}{P}}        &\text{select right choice}
      \\ \mid& \tm{\cpCase{x}{P}{Q}}    &\text{offer binary choice}
      &  \mid& \tm{\cpAbsurd{x}}        &\text{offer nullary choice}
    \end{array}
  \]
\end{definition}

@@

Reductions relate processes with their reduced forms.
\begin{definition}[Reduction]\label{def:cp-reduction}
  A reduction $\reducesto{P}{Q}$ denotes that the process $\tm{P}$ can reduce
  to the process $\tm{Q}$ in a single step.
  The relation $\Longrightarrow^\star$ is the reflexive, transitive closure of
  $\Longrightarrow$.
  Reductions can only be constructed as follows:
  \[
    \begin{array}{llcll}
      \cpRedAxCut1
      & \tm{\cpCut{x}{\cpLink{w}{x}}{P}}
      & \Longrightarrow
      & \tm{\cpSub{w}{x}{P}}
      \\
      \cpRedBetaTensParr
      & \tm{\cpCut{x}{\cpSend{x}{y}{P}{Q}}{\cpRecv{x}{y}{R}}}
      & \Longrightarrow
      & \tm{\cpCut{y}{P}{\cpCut{x}{Q}{R}}}
      \\
      \cpRedBetaOneBot
      & \tm{\cpCut{x}{\cpHalt{x}}{\cpWait{x}{P}}}
      & \Longrightarrow
      & \tm{P}
      \\
      \cpRedBetaPlusWith1
      & \tm{\cpCut{x}{\cpInl{x}{P}}{\cpCase{x}{Q}{R}}}
      & \Longrightarrow
      & \tm{\cpCut{x}{P}{Q}}
      \\
      \cpRedBetaPlusWith2
      & \tm{\cpCut{x}{\cpInr{x}{P}}{\cpCase{x}{Q}{R}}}
      & \Longrightarrow
      & \tm{\cpCut{x}{P}{R}}
      \\
      \cpRedKappaTens1
      & \tm{\cpCut{x}{\cpSend{y}{z}{P}{Q}}{R}}
      & \Longrightarrow \;
      & \tm{\cpSend{y}{z}{\cpCut{x}{P}{R}}{Q}}
        \quad \text{if} \; \notFreeIn{x}{Q}
      \\
      \cpRedKappaTens2
      & \tm{\cpCut{x}{\cpSend{y}{z}{P}{Q}}{R}}
      & \Longrightarrow \;
      & \tm{\cpSend{y}{z}{P}{\cpCut{x}{Q}{R}}}
        \quad \text{if} \; \notFreeIn{x}{P}
      \\
      \cpRedKappaParr
      & \tm{\cpCut{x}{\cpRecv{y}{z}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpRecv{y}{z}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaBot
      & \tm{\cpCut{x}{\cpWait{y}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpWait{y}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaPlus1
      & \tm{\cpCut{x}{\cpInl{y}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpInl{y}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaPlus2
      &\tm{\cpCut{x}{\cpInr{y}{P}}{R}}
      & \Longrightarrow \;
      & \tm{\cpInr{y}{\cpCut{x}{P}{R}}}
      \\
      \cpRedKappaWith
      & \tm{\cpCut{x}{\cpCase{y}{P}{Q}}{R}}
      & \Longrightarrow \;
      & \tm{\cpCase{y}{\cpCut{x}{P}{R}}{\cpCut{x}{Q}{R}}}
      \\
      \cpRedKappaTop
      & \tm{\cpCut{x}{\cpAbsurd{y}}{R}}
      & \Longrightarrow \;
      & \tm{\cpAbsurd{y}}
    \end{array}
  \]
  \begin{center}
    \begin{prooftree*}
      \AXC{$\reducesto{P}{P^\prime}$}
      \SYM{\cpRedGammaCut}
      \UIC{$\reducesto{\cpCut{x}{P}{Q}}{\cpCut{x}{P^\prime}{Q}}$}
    \end{prooftree*}
    \begin{prooftree*}
      \AXC{$\tm{P}\equiv\tm{Q}$}
      \AXC{$\reducesto{Q}{Q^\prime}$}
      \AXC{$\tm{Q^\prime}\equiv\tm{P^\prime}$}
      \SYM{\cpRedGammaEquiv}
      \TIC{$\reducesto{P}{P^\prime}$}
    \end{prooftree*}
  \end{center}
\end{definition}

Channels in \cp are typed using a session type system which corresponds to
classical linear logic.
A process communicating over a channel of type \ty{A \tens B} must be ready to
split into two independent processes (two processes which share no
channels), ready to communicate over a channel of type \ty{A} and \ty{B},
respectively, and process communicating over a channel of type \ty{A \parr B} is
dual to this, and therefore must be ready to start communicating concurrently
with two independent processes.
A process communicating over a channel of type \ty{A \plus B} can choose to
communicate either as \ty{A} or \ty{B}, and a process communicating over a
channel of type \ty{A \with B} must be able to communicate either as \ty{A} or
\ty{B}.
\begin{definition}[Types]\label{def:cp-types}
  \[
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
    \end{array}
  \]
\end{definition}
Duality plays a crucial role in both linear logic and session types.
In \cp, the two endpoints of a channel are assigned dual types.
This ensures that, for instance, whenever a process \emph{sends} across a
channel, the process on the other end of that channel is waiting to
\emph{receive}.
Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$.
Duality is an involutive function.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \begin{array}{lcllcllcllcl}
         \ty{(A \tens B)^\bot} &=& \ty{A^\bot \parr B^\bot}
      &  \ty{\one^\bot}        &=& \ty{\bot}
      &  \ty{(A \parr B)^\bot} &=& \ty{A^\bot \tens B^\bot}
      &  \ty{\bot^\bot}        &=& \ty{\one}
      \\ \ty{(A \plus B)^\bot} &=& \ty{A^\bot \with B^\bot}
      &  \ty{\nil^\bot}        &=& \ty{\top}
      &  \ty{(A \with B)^\bot} &=& \ty{A^\bot \plus B^\bot}
      &  \ty{\top^\bot}        &=& \ty{\nil}
    \end{array}
  \]
\end{definition}
Environments associate channels with types.
Names in environments must be unique, and environments $\ty{\Gamma}$ and
$\ty{\Delta}$ can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if
$\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} := \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}
  \) 
\end{definition}
Typing judgements associate processes with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$ denotes
  that the process \tm{P} communicates along channels $\tm{x_1}\dots\tm{x_n}$
  following protocols $\ty{A_1}\dots\ty{A_n}$.
  Typing judgements can be constructed using the following inference rules.
  \\[0.5\baselineskip]
  Structural Rules.
  \begin{center} \cpInfAx     \cpInfCut   \end{center}\vspace*{0.5\baselineskip}

  Logical Rules.
  \begin{center} \cpInfTens   \cpInfParr  \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfOne    \cpInfBot   \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfPlus1  \cpInfPlus2 \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfWith               \end{center}\vspace*{0.5\baselineskip}
  \begin{center} \cpInfNil    \cpInfTop   \end{center}\vspace*{0.5\baselineskip}
\end{definition}
\cp enjoys subject reduction, progress, and termination~\cite{wadler2012}.

@@

\begin{lemma}[Type preservation for \equiv]\label{lem:hcp-preservation-equiv}
  If $\tm{P}\equiv\tm{Q}$, then $\seq[P]{\mathcal{G}}$ iff $\seq[Q]{\mathcal G}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\tm{P}\equiv\tm{Q}$.
\end{proof}

\begin{theorem}[Subject reduction]\label{thm:hcp-preservation}
  If $\seq[P]{\mathcal{G}}$ and $\reducesto{P}{Q}$, then $\seq[Q]{\mathcal G}$.
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reducesto{P}{Q}$.
\end{proof}
